---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

devtools::load_all()
pkgs <- simulatingMeta:::.get_packages()
```

# Understanding meta-analysis through data simulation with applications to power analysis

<!-- badges: start -->
[<img alt="alt_text" src="https://img.shields.io/badge/OSF-https%3A%2F%2Fosf.io%2F54djn%2F-blue" />](https://osf.io/54djn/)
<!-- badges: end -->

This repository contains the code to reproduce the analysis of the paper **Understanding meta-analysis through data simulation with applications to power analysis** by Gambarota Filippo and Gianmarco Altoè. The project can also be found on the Open Science Framework https://osf.io/54djn/ with a preprint uploaded on PsyArXiv [https://psyarxiv.com/br6vy/](https://psyarxiv.com/br6vy/).
  
## Usage

Functions used in the paper and supplementary materials are documented and can be used opening the `simulating-meta-analysis.Rproj` file and using the `devtools::load_all()` function (this requires the `devtools` package installed). Then each function will be available as using a standard R package.

## Documents

The repository (`documents/` folder) contains the paper and supplementary materials created using R Markdown and the [`papaja`](https://cran.r-project.org/web/packages/papaja/index.html) package.

| **Document** | **Rendered** | **Source** |
|---|---|---|
| Paper | [pdf](documents/paper/paper.pdf) | [rmd](documents/paper/paper.Rmd) |
| Supplementary Materials | [pdf](documents/paper/supplementary.pdf) | [rmd](documents/paper/supplementary.Rmd) |

## Repository

The project is organized as follows:

- `R`: contains the simulation and utility functions used in the paper and the supplementary materials
- `documents`: contains the `.Rmd` files to reproduce the paper and supplementary materials manuscripts
- `scripts`: contains the script used to produce complex objects and figures
  - `scripts\r-img.R`: create the paper figures. Figures that are not created in this script using `ggplot2` are created with the open-source program [*Inkscape*](https://inkscape.org/) and contained in the `../img/` folders.
- `objects`: contains the result of objects created by the `scripts/` scripts.

## Testing

The custom functions are tested within the `scripts/test.R` script. The idea is to use the same simulation codes of the paper increasing the $k$ and $n$ values and check the recovery of simulated parameters using the `check_sim()` function.

# Session

## Session info

```{r, echo=FALSE}
sinfo <- sessioninfo::session_info()$platform
sinfo <- sinfo[names(sinfo) != "date"]
paste0(
  cli::style_bold(cli::col_blue(names(sinfo))), ": ", sinfo
) |> 
  cat(sep = "\n")
```

## Packages

```{r, echo = FALSE, message=FALSE, warning=FALSE, results='asis'}
sprintf("- `%s` (%s)", pkgs$pkg, pkgs$version) |> 
  cat(sep = "\n")
```